---
title: "Code to calculate impact score"
author: EICAT working group
date: 2023
output: pdf_document
---






The code below uses a specific data structure, from the Acacia data of Sabrina.

```{r}

library(plyr)
library(dplyr)
library(readxl)

dat <- readxl::read_xls('EICAT_Acacia_dataset.xls')

dat$ordination_score <- as.numeric(dat$ordination_score)
dat$ordination_confidence <- as.numeric(dat$ordination_confidence)
# must rename some columns that have odd formatting issues
# this assumes that column 2 is the alienSpecies column
colnames(dat)[2] <- 'alienSpecies'


# estimated weighted confidence (done below in the `makePlot` function)
conf <- dat |> 
    group_by(alienSpecies) |>
    summarise(weightMean=weighted.mean(ordination_score, w=ordination_confidence, na.rm=TRUE), 
    numRecords=length(!is.na(ordination_score)))
conf2 <- conf[which(!is.nan(conf$weightMean)), ]

unSpecies <- conf2$alienSpecies

# an example species for initial exploration
tmp <- dplyr::filter(dat, alienSpecies=='Acacia dealbata')

```



```{r}

#' @param dat the data object in similar format to the acacia 
#'  columns named ordination_score, ordination_confidence, and alienSpecies
#' @param species species name
#' @param save do you want the plot to be saved as png?
#' @param type type=1 is original bit, type=2 is second pass
#'
#' @results outputs weighted mean and plot

makePlot <- function(dat, species='Acacia dealbata', 
    save=FALSE, type=1, axis=TRUE){
    tmp <- dat[which(dat$alienSpecies == species), ]

    Wmeansd <- function(x, weights) {
        xm <- weighted.mean(x, weights, na.rm = TRUE)
        variance <- sum((weights * (x - xm)^2) / (sum(weights[!is.na(x)]) - 1), 
            na.rm = TRUE)
        sd <- sqrt(variance)
        return(c(xm,sd))
    }
    score <- try(Wmeansd(tmp$ordination_score, 
        w=tmp$ordination_confidence))
    if(inherits(score, 'try-error')){
        score <- c(NA, NA)
    }

    species <- unlist(as.character(species))
    if(save){png('tad.png', width=6, height=3)}
    plot(x=seq(0,1, length.out=10), y=rep(0.5,10), type='l', 
        axes=FALSE, xlab='', ylab='', 
        main=substitute(paste(italic(species))), lwd=2,
        ylim=c(0.45,0.55))

    if(type==1){
        points(x=tmp$ordination_score, y=jitter(rep(0.5, nrow(tmp))), 
            pch=16, col=viridis::viridis(5)[as.factor(cut(tmp$ordination_confidence, 5))])
    }
    if(type==2){
        tmp2 <- (tmp[, c('ordination_score', 'ordination_confidence')])
        tmp2 <- tmp[!is.na(tmp2$ordination_score), ]
        tmp2 <- tmp2[order(tmp2$ordination_score), ]
        y2 <- table(tmp2$ordination_score)
        pos <- seq(0.5, 0.55, length.out=max(y2))
        y3 <- c()
        #colz <- viridis::viridis(3)[as.factor(cut(tmp2$ordination_confidence, 3))]
        for(i in 1:length(y2)){
            y3 <- c(y3, pos[1:y2[i]])
            #colz <- c(colz, colz[1:y2[i]])
        }
        points(x=tmp2$ordination_score, 
            y=y3, 
            pch=16, col=viridis::viridis(3)[as.factor(cut(tmp2$ordination_confidence, 3))])

    }
    if(axis){
        axis(1, at=c(0, 0.25, 0.5, 0.75, 1), 
            labels=c('Minimal', 'Minor', 'Moderate', 'Major', 'Massive'))
    }
    segments(x0=score[1]+score[2], x1=score[1]-score[2], 
        y0=0.5, lwd=4, col='grey')
    points(x=score[1], y=0.5, pch=16, cex=4)
    text(x=0.5, y=0.48, 'Impact score')
    if(save){dev.off()}

    return(score)
}

```





Visualizing impact score for each Acacia species for which we had enough data. Outputs the weighted mean and weighted standard deviation for each and creates the plots. 

```{r}

pdf('figures/leAcacia.pdf')
layout(matrix(1:15, ncol=3))
par(mar=c(1,1,1,0))
for(i in 1:15){
    makePlot(dat, species=conf2$alienSpecies[i], type=2, axis=FALSE)
}
legend('bottomright', c('Low', 'Moderate', 'High'), pch=16, ncol=3, 
    bty='n', col=viridis::viridis(3), title='Confidence')

dev.off()

```






### Pulling out mechanism 

I like the idea of using the underlying mechanism, so the code below pulls out the mechanism and uses a binary column to indicate proposed mechanism for each record. 

```{r}

indz <- regmatches(dat$Mechanism, 
    gregexpr("[[:digit:]]+", dat$Mechanism))  

unMechz <- unique(unlist(indz))
mechz <- matrix(0, ncol=length(unMechz), nrow=nrow(dat))
colnames(mechz) <- unMechz


for(i in 1:length(inds)){
    if(length(indz[[i]])!=0 ){
        mechz[i, colnames(mechz) %in% indz[[i]]] <- 1
    }
}

mechz <- as.data.frame(mechz)

dat2 <- as.data.frame(cbind(dat, mechz))

```




Some plotting of the distribution of the number of mechanisms per record and the relationship between the impact score (not the weighted one) and the number of mechanisms. 

```{r}
png('figures/distroMechz.png')
hist(rowSums(dat2[,17:22]), 
    xlab='Number of mechanisms', main='', 
    col=adjustcolor('dodgerblue', 0.5), border='dodgerblue')
dev.off()



png('figures/scoreAndMechNum.png')
plot(as.factor(dat2$ordination_score), rowSums(dat2[,17:22]), 
    pch=16, xlab='Impact score', ylab='Number of mechanisms')
dev.off()

```




Starting to visualize the distribution of mechanisms per record per species

```{r}

tmp2 <- dplyr::filter(dat2, alienSpecies=='Acacia dealbata')

tmp3 <- as.matrix(tmp2[,17:23])
tmp3[tmp3==0] <- NA 
lattice::levelplot(tmp3)

```

